---
title: "IJohnson Mod04 QFISH Assignment"
author: "Ivanya Johnson"
date: "20-09-2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QFish Shark Control Program Data

To download data go to:
https://qfish.fisheries.qld.gov.au/query/number-caught-by-area-calendar-year-and-species-group/table


## Install and Load Packages

Be sure to install the following packages if they are not already installed.
```{r Packages}
# install.packages("janitor")

library(janitor)
library(tidyverse)
library(readr)
library(ggplot2)
library(ggthemes)
```


## Load Data
```{r}
totalcatch <- # Name data frame
  read_csv("Shark Catch.csv") |> # Select data to import
  row_to_names(row_number = 2, remove_rows_above = FALSE) |> # Select the row to use for variable names of data frame.
  clean_names() # Creates unique names for the variables. If not included, R will give variables with common names the same name

head(totalcatch) # Look at data

totalcatch <- subset(totalcatch, select = -(na_2)) # Removing empty column

summary(totalcatch) # Summary of the data
```


## Tidying Data

Now, we need to separate the data into individual data sets. These data sets will be for the categories of species (sharks, mammals, turtles, other).
```{r}
# Shark Catch
sharks <- # Creates a new data frame titled sharks
  totalcatch |>
  select(area | contains("shark")) |>
  row_to_names(row_number = 1) |>
  rename(area = 1) |>
  mutate(across(-c(area), as.numeric))

# Turtle Catch
turtles <-
  totalcatch |>
  select(area | contains("turtle")) |> # Select the column titled area and any column with the word "turtle"
  row_to_names(row_number = 1) |> # This function elevates the selected row (row = 1) to become the new column names
  rename(area = 1) |> 
  mutate(across(-c(area), as.numeric)) # Mutates the columns selected to input data values and make them numeric

# Mammal Catch
mammals <-
  totalcatch |>
  select(area | contains("mammal")) |>
  row_to_names(row_number = 1) |>
  rename(area = 1) |>
  mutate(across(-c(area), as.numeric))

# Others Catch
other <-
  totalcatch |>
  select(area | contains("other")) |>
  row_to_names(row_number = 1) |>
  rename(area = 1) |>
  mutate(across(-c(area), as.numeric))

head(sharks)
head(turtles)
head(mammals)
head(other)
```


## Pivot Data Tables

The tables we created for each group is called a wide table. We want to pivot them longer so we have two columns labeled year and count.
```{r}
catch_sharks <-
  sharks |>
  pivot_longer(!area, names_to = "Year",
               values_to = "Count")

catch_turtles <-
  turtles |>
  pivot_longer(!area, names_to = "Year",
               values_to = "Count")

catch_mammals <-
  mammals |>
  pivot_longer(!area, names_to = "Year",
               values_to = "Count")

catch_other <-
  other |>
  pivot_longer(!area, names_to = "Year",
               values_to = "Count")
```


## Making Stats Tibbles

You don't HAVE to do this, but it helps to filter out any unwanted variables, columns, or rows. It also makes it easier to combine the data together.
```{r}
sharks_stats <- 
catch_sharks |>
  filter(area != "Grand Total") |> # Filters out all the 'Grand Total' rows under the area column
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE), # Calculate the mean of the values in the row called Count
            SD = sd(Count, na.rm = TRUE)) |> # Calculates the standard deviation
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Sharks")

turtles_stats <- 
catch_turtles |>
  filter(area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Turtles")

mammals_stats <- 
catch_mammals |>
  filter(area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Mammals")

other_stats <- 
catch_other |>
  filter(area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Other")
```


## Combine the new statistical tables

Now it's time to combine the new stats into one!
```{r}
Total_stats <- 
  bind_rows(sharks_stats, turtles_stats,
            mammals_stats, other_stats)
```

## Plot Data

Now that we have tidy'd, subset, and combined our data we can make plots! The first plot we will make is a line chart showcasing the mean capture of each taxa.
```{r}
ggplot(data = Total_stats, aes(x =Year, y = Mean, fill = Taxa, shape = Taxa), color = "black") +
  geom_line() +
  geom_pointrange(aes(ymin = Mean - SD, ymax = Mean + SD)) +
    scale_shape_manual(values = c(21, 22, 23, 24))
```