---
title: "MB5370 Module 04. Data Science in R 4"
author: "Ivanya Johnson"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 4: Spatial Data in R


## Installing / Loading Spatial Packages
I have already installed these packages so I have hashtagged the installation code out.
```{r}
# install.packages("sf")
# install.packages("terra")
# install.packages("tmap")

library(tidyverse)

library(ggplot2)

library(sf) # simple features

library(terra) # raster

library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized

library(readr)
```


## Importing Copepod Data
```{r}
dat <- read_csv("../data-for-course/copepods_raw.csv") # Because I am in a different working directory, I need to tell R to go back and find the folder titled data-for-course to find the copepods csv

dat
```


## Data Exploration

### Check Coordinates
```{r}
ggplot(dat) +
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

Add the stat_smooth function
```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) +
  stat_smooth() +
  geom_point()
```


### Get Going With Maps
```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"),
                 crs = 4326)
```


```{r}
?st_as_sf
```
st_as_sf converts different data types to simple features
  - coords gives the names of the columns that relate to the spatial coordinates (in order of X coordinate followed by Y coordinate)


### Coordinate Reference System

  - crs stands for 'coordinate reference system'
  - CRS are required for 2D mapping to compensate for the lumpy, spherical (3D) nature of the earth

```{r}
crs4326 <- st_crs(4326)

crs4326 # Look at the whole crs

crs4326$Name

crs4326$wkt
```


### Feature Collection (Points)
```{r}
sdat
```


### Cartography
```{r}
plot(sdat["richness_raw"])

plot(sdat)
```


## Thematic Maps for Communication

### Using tmap
  - tm_dots to plot dots of the coordinates. Other options are tm_polygons and tm_symbols

```{r}
tm_shape(sdat) +
  tm_dots(col = "richness_raw")
```


Use tmap_save to save the map to your working directory
```{r}
tmap_save(filename = "Richness-map.png",
          width = 600, height = 600)
```


## Mapping Spatial Polygons as Layers

### Loading Shapefiles
Shapefiles are not ideal as they're inefficient at storing data and to save 1 shapefile you create multiple files. This means, a bit of the file might be lost if you transfer the data somewhere else.

```{r}
aus <- st_read("../data-for-course/spatial-data/Aussie/Aussie.shp")
shelf <- st_read("../data-for-course/spatial-data/aus_shelf/aus_shelf.shp")

aus
shelf
```


### Mapping Polygons

Blank map of Australia
```{r}
tm_shape(shelf) +
  tm_polygons()
```

With this code, we see the data points
```{r}
tm_shape(shelf, bbox = sdat) +
  tm_polygons() +
  tm_shape(aus) +
  tm_polygons() +
  tm_shape(sdat) +
  tm_dots()
```


## Exploring tmap

Follow the instructions in the Help Window
```{r}
vignette('tmap-getstarted')
```

### Create a Map of the world
The object World is a spatial object of class sf from the sf package; it is a data.frame with a special column that contains a geometry for each row, in this case polygons. In order to plot it in tmap, you first need to specify it with tm_shape. Layers can be added with the + operator, in this case tm_polygons. There are many layer functions in tmap, which can easily be found in the documentation by their tm_ prefix. See also ?'tmap-element'.

```{r}
data("World")

tm_shape(World) +
  tm_polygons("HPI")
```

### Interactive Maps
Each map can be plotted as a static image or viewed interactively using "plot" and "view" modes, respectively. The mode can be set with the function tmap_mode, and toggling between the modes can be done with the ‘switch’ ttm() (which stands for toggle thematic map.

```{r}
tmap_mode("view")

tm_shape(World) +
  tm_polygons("HPI")
```

### Multiple shapes and layers
A shape is a spatial object (with a class from sf, sp, stars, or raster). Multiple shapes and also multiple layers per shape can be plotted.

```{r}
data(World, metro, rivers, land)

tmap_mode("plot")

tm_shape(land) + 
  tm_raster("elevation", palette = terrain.colors(10)) +
  tm_shape(World) +
  tm_borders("white", lwd = .5) +
  tm_text("iso_a3", size = "AREA") +
  tm_shape(metro) +
  tm_symbols(col = "red", size = "pop2020", scale = .5) +
  tm_legend(show = FALSE)
```

### Facets
Facets can be created in three ways:

  1. By assigning multiple variable names to one aesthetic (in this example the first argument of tm_polygons:
```{r}
tmap_mode("view")

tm_shape(World) +
  tm_polygons(c("HPI", "economy")) +
  tm_facets(sync = TRUE, ncol = 2)
```


  2. By splitting the spatial data with the by argument of tm_facets:
```{r}
tmap_mode("plot")

data(NLD_muni)

NLD_muni$perc_men <- NLD_muni$pop_men / NLD_muni$population * 100

tm_shape(NLD_muni) +
  tm_polygons("perc_men", palette = "RdYlBu") +
  tm_facets(by = "province")
```


  3. By using the tmap_arrange function:
```{r}
tmap_mode("plot")

data(NLD_muni)

tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)

tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```


### Basemaps and Overlay Tile Maps
Tiled basemaps can be added with the layer function tm_basemap. Semi-transparent overlay maps (for example annotation labels) can be added with tm_tiles.

```{r}
tmap_mode("view")

tm_basemap("Stamen.Watercolor") +
  tm_shape(metro) + tm_bubbles(size = "pop2020", col = "red") +
  tm_tiles("Stamen.TonerLabels")
```

Note: For some reason, the above code does not produce the map. it only produces the red bubbles!


### Options and Styles
The functions tm_layout and tm_view are used to specify the map layout and the interactive aspects respectively. These functions can be used in the same way as the layer functions.

```{r}
tmap_mode("plot")

tm_shape(World) +
  tm_polygons("HPI") +
  tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```


These options, as well as a couple of others, can also be set within with tmap_options, which works in the same way as the base R function options. The main advantage is that these options are set globally, so they do not have to be specified in each map, for the duration of the session.

```{r}
tmap_options(bg.color = "black", legend.text.color = "white")

tm_shape(World) +
  tm_polygons("HPI", legend.title = "Happy Planet Index")
```


A style is a certain configuration of the tmap options.
```{r}
tmap_style("classic")

tm_shape(World) +
  tm_polygons("HPI", legend.title = "Happy Planet Index")

tmap_options_diff()

tmap_options_reset()
```


## Exporting Maps
The following is the code used to save the last map created

```{r}
?tmap_save

tm <- tm_shape(World) +
    tm_polygons("HPI", legend.title = "Happy Planet Index")

## save an image ("plot" mode)
tmap_save(tm, filename = "world_map.png")

## save as stand-alone HTML file ("view" mode)
tmap_save(tm, filename = "world_map.html")
```