---
title: "MB5370 Module 04. Data Science in R 1"
author: "Ivanya Johnson"
output: html_document
date: "2024-08-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# WORKSHOP 1: Reproducible Research & Version Control


## Install and load tidyverse package

These first lines of code include the package needed to use Git in RStudio and how we set our credentials.
```{r}
# install.packages("devtools")
# devtools::install_github("r-lib/conflicted")
# gitcreds::gitcreds_set()
```


## Install and Load packages and Debugging

Installing and loading the tidyverse packages. I set my preferences to use dplyr functions over others and loaded any other packages that come with ggplot and tidyverse.
```{r}
# install.packages("tidyverse")

library(tidyverse)
library(conflicted)
library(dplyr)

conflicts_prefer(dplyr::filter) # Use to set preferences
conflicts_prefer(dplyr::lag) # Use to set preferences 

library(ggthemes)

library(ggplot2)
```


## Load the data (mpg is built into ggplot2)

I loaded the Palmer Penguin data as a backup. However, for this workshop, I used the mpg data provided in ggplot.
```{r}
data(mpg)

library(palmerpenguins) #Palmer Penguins Data
```


## Quick data checks

Look at the data to see what we're working with, columns, rows, etc.
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


## Create first plot

When you’re creating a plot (using ggplot), you essentially need two attributes of a plot: a geom and aesthetics.Using ggplot to create a point plot of the mpg data.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# displ is the engine size and hwy is the fuel efficiency
# geom_point adds a layer of points to the plot
```
The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel.



## Graphing Template

```{r}
ggplot(data = mpg) +
  geom_line(mapping = aes(x = displ, y = hwy))
```
This line graph doesn't look very neat. It's scattered and all over the place. There are a few ways to clean it up!



## Troubleshooting

### Aesthetic Mappings
You can add a third variable to plots, such as class

#### Change point color by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

#### Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!"Using size for a discrete variable is not advised".


#### Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Another warning!! "Using alpha for a discrete variable is not advised".


#### Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```
You must specify shapes manually because the shape function can only handle a max of 6 values


#### Make all points purple
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "purple")
```


#### Change point color based on TRUE or FALSE
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```
Makes it a logical function (True or False) and changes color to reflect that


## Faceting and Panel Plots

### facet_wrap
Use facet_wrap to create a subset of your data; only use for discrete variables.

#### Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


### facet_grid
Use facet_grid to do this with more than one variable.

#### A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


### facet_grid by just row (or column)
Use a . if you don't want to facet in the rows or column dimension

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)

  #facet_grid(. ~ cyl) # Alternatively

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

## Exercise 1:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

```{r}
?facet_wrap # Pulls up the Help tab explaining the arguments of this function and what it is used for.
```
nrows is the number of rows, ncol is the number of columns, shrink is used with scales. If it's true that the scale needs to be fixed, it will shrink the output of statistics. Axes determines which axes will be drawn in the case of fixed scales. Margins is the default for axes.



## Fitting Simple Lines

We already used points above.

However, ggplot2 can use a variety of geom objects to represent the data. We might want to use bar plots, line charts, box plots and so on. We can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

### To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 


### Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
# Can also use "lty = " for the same result
```

To set the lines to different colors, you must input the variable you want the colors to be based on


## Complex Plots in ggplot

### Change line color based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```


### Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
With the above code, the top and second rows are duplicates so if you wanted to change the x variable you'd have to do it in multiple locations


A better way... you can make global mappings so you'd only have to change it in one location.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```
By doing this, you can change certain features of one set without impacting the other.Now we have different colors to indicate the raw data (geom_point) and one color for the overall relationship (geom_smooth). Neat!



### Specifying different data for each layer using a filter called "subcompact" to select a subset of the data and plot ONLY that subset
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```


## Exercise 2:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

- Use geom_line for a line chart, geom_boxplot for a box plot, geom_histogram for a histogram, and geom_area for an area plot.


2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

- Prediction: The code will give us the same plot from earlier (right above Exercise 2 in Multiple geoms). One chart using both point data and a smooth line (kinda like a line of best fit).
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```


3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
The above code give the same result because, as we saw earlier, you can duplicate the data or make it global. In the bottom code, we have simply duplicated the data in the top code.

## Transformations and stats

### Load diamond data
```{r}
data("diamonds")
glimpse(diamonds)
```

### Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

# You can also use 'stat_count' and it gives the same plot
```

### Bar plot using stat_count
```{r}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```


## Overriding defaults in ggplot2
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```


### What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```
If you get the following:
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
You can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))


### Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


## Aesthetic Adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like color or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colors.

### Adjustment 1
Has grey bars with a highlighted color around each bar
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```


### Adjustment 2
Changes the color for each bar
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```


Question: Does anyone notice anything different in the color scheme of this plot?
(Hint: It's in the viridis color palette (color blind friendly), but why is it different from the color palette we used earlier?)
Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


### Filling by a variable
Now try using these aesthetics to color by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.

Stacks the bars and has a different color for each clarity
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


### To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


### To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


### Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


### position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

### Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatter plots but not bar plots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```
