---
title: "MB5370 Module 04. Data Science in R 2"
author: "Ivanya Johnson"
date: "2024-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# WORKSHOP 2: Using ggplot2 for Communication


## Load packages

Loading the tidyverse and ggplot2 packages. I set my preferences to use dplyr functions over others and loaded any other packages that come with ggplot and tidyverse.
```{r}
library(tidyverse)
library(conflicted)
library(dplyr)

conflicts_prefer(dplyr::filter) # Use to set preferences
conflicts_prefer(dplyr::lag) # Use to set preferences 

library(ggthemes)

library(ggplot2)
```


## Labels

```{r}
data(mpg)

library(palmerpenguins) #Palmer Penguins Data
```


### Annotations

```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


### Scales

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# displ is the engine size and hwy is the fuel efficiency
# geom_point adds a layer of points to the plot
```


### Axis Ticks

```{r}
ggplot(data = mpg) +
  geom_line(mapping = aes(x = displ, y = hwy))
```


### Legends and Color Schemes

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```


### Replacing a Scale

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "purple")
```


### Popular Color Package

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```



### Themes

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


#### Theme that Ben wrote himself so his plots are the same style. We can do this to format our own as well

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```


### Saving / Exporting Plots

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```
